---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
My project1 for the Reproducible Research coursera course through Johns Hopkins University.

R Markdown Websites used: 

- [R Markdown - Dynamic Documents for R](http://rmarkdown.rstudio.com)
- [Using R Markdown](https://support.rstudio.com/hc/en-us/articles/200552086-Using-R-Markdown)
- [knitr](http://yihui.name/knitr/)

## Loading and preprocessing the data
```{r read_data_file, echo = TRUE}
library(data.table)
data<-read.csv("activity.csv", header=TRUE)
DT<-data.table(data)
```
**Note: I'm using the data table package for this project.**  
[Data Tables](http://cran.r-project.org/web/packages/data.table/index.html)

## What is mean total number of steps taken per day?
```{r mean_hist, echo=TRUE}
groupByDate<-DT[, sum(steps), by = date]
setnames(groupByDate,"V1", "steps")
hist(groupByDate[,steps], main="Histogram of Steps Per Day", 
     xlab="Number of Steps", ylab="Number of Days",
     ylim=c(0,30))
mean_val<-mean(groupByDate$steps, na.rm=TRUE)
median_val<-median(groupByDate$steps, na.rm = TRUE)
```
***The mean value is `r formatC(mean_val, digits=2, format="f")`.***  
***The median value is `r median_val`.***

## What is the average daily activity pattern?
```{r dailey_activity_pattern, echo=TRUE}
groupByTime<-DT[!is.na(steps), mean(steps), by=interval]
setnames(groupByTime,"V1", "steps")
plot(groupByTime[,interval], groupByTime[,steps], type="l",
     xlab="Interval", ylab="Steps", main="Daily Activity Pattern")
temp<-groupByTime[order(steps, decreasing = TRUE)]
mostSteps<-temp[1,interval]

```
***The interval with the most steps is `r mostSteps`.***


## Imputing missing values
```{r na_count, echo=TRUE}
naCount<-sum(DT[,is.na(steps)])
```
The total number of NA values is **`r naCount`**.  
***Impute Stragegy: NA values replaced with the average steps for the interval.***
```{r impute_values, echo=TRUE}
DT[, steps:=as.numeric(steps)][is.na(steps),steps:={tmp<-interval; 
                                                    groupByTime[interval==tmp,steps]}]
groupByDate<-DT[, sum(steps), by = date]
setnames(groupByDate,"V1", "steps")
hist(groupByDate[,steps], main="Histogram of Steps Per Day", 
     xlab="Number of Steps", ylab="Number of Days",
     ylim=c(0,40))
mean_val<-mean(groupByDate$steps)
median_val<-median(groupByDate$steps)
```
The mean value using imputed data is **`r formatC(mean_val, digits=2, format="f")`**.  
The median value using imputed data is **`r formatC(median_val, digits=0, format="f")`**.  
***There is very little difference between using the raw data set (with NA values) and the imputed data set***



## Are there differences in activity patterns between weekdays and weekends?
```{r weekend, echo=TRUE}
library(ggplot2)
library(gridExtra)
DT[, date:=as.Date(date)]
str(DT) #date field now of type Date instead of factor
DT[, dayOfWeek:=ifelse(weekdays(date) %in% c("Saturday", "Sunday"), "weekend", "weekday")]
weekdayGroup<-DT[dayOfWeek=="weekday"][, mean(steps), by=interval]
weekendGroup<-DT[dayOfWeek=="weekend"][, mean(steps), by=interval]
setnames(weekdayGroup,"V1", "steps")
setnames(weekendGroup,"V1", "steps")
p1<-ggplot(weekdayGroup,aes(x=interval, y=steps)) + geom_line() + 
    labs(title = "Weekdays", y="Number of Steps", x="Five Minute Intervals") +
    scale_y_continuous(limits=c(0,250))
p2<-ggplot(weekendGroup,aes(x=interval, y=steps)) + geom_line() + 
    labs(title = "Weekends", y="Number of Steps", x="Five Minute Intervals") +
    scale_y_continuous(limits=c(0,250))
grid.arrange(p1, p2, ncol=1)
```
  
***There are differences in the patterns between the weekdays and the weekends. During the weekdays the majority of the steps are in the morning. On the weekends the steps are more evenly distributed through out the day.***


